---
title: "User-defined functions and control flow (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA)
pulse <- read.table("../data/pulse.txt") 
survey <- read.delim("../data/survey.txt") 
source( "../R_course_setup.R" )
```

```{r, echo=FALSE, eval=FALSE}
# |classification|bmi|  
# |---------|--------|
# |underweight|<18.5|
# |normal|18.5-24.9|
# |overweight| 25-29.9|
# |obese| >29.9|

```

# User-defined functions and control flow (tasks)

## Tasks

### functions (0) {#functions0}

- Write the function `even` which tests whether a number is even.

```{r}
# (v1)
even <- function(x) {
  x %% 2 == 0
}
```

- Write an alternative version of function `even` using `odd` function.


```{r}
# (v2)
even <- function(x) {
  !odd(x)
}

```

- Write a function, given values `x` and `m`, that tests whether `x` is integer divisible by `m`.

```{r}
divisible <- function(x,m) {
  x %% m == 0
}
```



- There are two systems for weight and height measurements, metric (cm,kg) and imperial (inch,lbs). The relation between the measurements in different systems is  (1 kg = 2.2 lbs) and  (1 inch = 2.54 cm). Write the following unit conversion functions:
    - cm2meter : convert cm to meter
    - lbs2kg   : convert weight unit pound to kg
    - inch2cm  : convert inch to cm

```{r, eval=FALSE, echo=TRUE}
# examples:
> inch2cm(70)
[1] 177.8
> lbs2kg(10)
[1] 4.545455
> cm2m(190)
[1] 1.9
```

```{r}
cm2m    <- function(x) x*0.01  # cm to meter
lbs2kg  <- function(x) x/2.2   # pound to kg
inch2cm <- function(x) x*2.54  # inch to cm
```

- (*) Generate the following sequence:

```{r, echo=FALSE, eval=TRUE}
divisible <- function(x,m) {
  x %% m == 0
}
v <- 1:100
v[sapply(v,function(x) divisible(x,7))]
```

```{r}
v <- 1:100
v[sapply(v,function(x) divisible(x,7))]
```


### functions (1) {#functions1}


  - The following table is a BMI classification. Write a function to determine 
the classification given a BMI value.

<center>
```{r, eval=TRUE, echo=FALSE}
library("htmlTable")
htmlTable( data.frame(classification=c("underweight","normal","overweight","obese"),
           bmi=c("<18.5","18.5-24.9","25-29.9","\\>30"), stringsAsFactors = F),  col.rgroup = c("none", "#F7F7F7") )
```
</center>



```{r, eval=FALSE, echo=TRUE}
bmiClass <- function(bmi) {
  # Clissifies BMI value given as argument into categories:
  #
  #         {"underweight","normal","overweight","obese"}
  #
  # your code: 
  ...
}  
# examples:
> bmiClass(19)
[1] "normal"
> bmiClass(29.9)
[1] "overweight"
> bmiClass(30)
[1] "obese"
```


```{r}
#
# bmiClass 
#
bmiClass <- function(bmi) {
  # Classifies BMI value given as argument into categories {"underweight","normal","overweight","obese"}
  #
  if (bmi < 18.5) {
    classification <- "underweight"
  } else if(18.5 <= bmi & bmi <= 24.9) {
    classification <- "normal"
  } else if(25 <= bmi & bmi <= 29.9) {
    classification <- "overweight"
  } else {
    classification <- "obese"
  }  
  classification
}
```

  
  
  

### functions (2)  {#functions}

  - Write a function calculating BMI with arguments weight(kg) and height(cm). BMI in the metric system is defined as weight/(height^2) with weight and height in units of kilogram (kg) and meter (m). Use `cm2m` function from previous exercise.   

```{r}

bmi <- function(weight,height) {
  # Calculates the BMI given weight (kg) and height (cm). 
  # weight : kg unit as value
  # height : cm unit as value
  weight/( cm2m( height )^2 )
}
```

  - Upgrade the functions `lbs2kg` and `inch2cm` as such that the conversions can be applied in the opposite direction. You can do this by adding an additional logical argument `invert`  to the function with the default value set to FALSE. For example a call like:
  
  
```{r, eval=FALSE, echo=TRUE}
lbs2kg(65, invert=TRUE) # convert kg to lbs
```

will reverse the conversion from kg to lbs. 


```{r}
lbs2kg  <- function(x, invert=FALSE) 
  if (!invert) {
    x/2.2   # pound to kg
  } else {
    x*2.2   # kg to pound
  }

inch2cm  <- function(x, invert=FALSE) 
  if (!invert) {
    x*2.54  # inch to cm
  } else {
    x/2.54   # cm to inch
  }

```
  
  - (*) Update BMI function with an additional logical argument `imperial` with the default value set to FALSE. The function will accept both systems of measurements (metric and imperial). Use the `imperial` argument in the function as a condition to convert imperial measurement to metric when the input values are in imperial. The function can be called  with values for `weight` and `height` in metric system (kg,cm) or imperial (lbs,inch).

  
```{r}

bmi <- function(weight,height,imperial=FALSE) {
  # Calculates the BMI given weight and height in either metric or imperial 
  # systems. By default the metric system is assumed (imperial=FALSE). 
  # 
  # weight : pound or kg units as values
  # height : inch or cm units as values
  #
  if (imperial) {
    height <- inch2cm(height)
    weight <- lbs2kg(weight)
  }
  weight/( cm2m( height )^2 )
}
```


### apply {#apply}

  - Calculate the mean age and height in survey data using apply function. Note that there are missing data (hint: see na.rm argument to mean). 
  
```{r}
# Two solutions is given.

# (1) anonymous function calls mean
apply( survey[,c("age","height")], 2, function(x)  mean(x,na.rm=TRUE) )

# (2) direct call to mean with argument `na.rm=TRUE` is passed to mean function through  apply ... argument (see ?apply)
apply( survey[,c("age","height")], 2, mean, na.rm=TRUE)

```

  - In the previous exercise we removed the missing values per variable. A more fair solution would be to remove the observation completely where one or both variables are missing. Hint : ?complete.cases
  
  
```{r}
# using an intermediate variable 'd' 
d <- survey[,c("age","height")] 
d <- d[complete.cases(d),] # update d where all observations are complete cases (contain no NA's)
apply(d,2, mean) 

```

  
  - (*) Upgrade bmiClass function in the previous exercises to process a vector of BMI values. 
 
```{r}

bmiClass <- function(bmi) {
  # Given a 'vector' of BMI values as argument return the corresponding BMI classifications.
  #
  bmiClassFunc <- function(bmi) {  
    # Clissifies a BMI value given as argument into categories {"underweight","normal",
    # "overweight","obese"}
    #
    if (bmi < 18.5) {
      classification <- "underweight"
    } else if(18.5 <= bmi & bmi <= 24.9) {
      classification <- "normal"
    } else if(25 <= bmi & bmi <= 29.9) {
      classification <- "overweight"
    } else {
      classification <- "obese"
    }  
    classification
  }
  sapply(bmi,bmiClassFunc)
}

```


### function argument check {#argcheck}

```{r}
bmiClass <- function(bmi) {
  # Clissifies BMI value given as argument into categories {"underweight","normal","overweight","obese"}
  #
  if (!is.numeric(bmi))
    stop("numeric type expected !")
  if (bmi < 18.5) {
    classification <- "underweight"
  } else if(18.5 <= bmi & bmi <= 24.9) {
    classification <- "normal"
  } else if(25 <= bmi & bmi <= 29.9) {
    classification <- "overweight"
  } else {
    classification <- "obese"
  }  
  classification
}
```


  - Update the function bmiClass as such to error on wrong input type, e.g.

```{r, echo=TRUE,eval=TRUE, error=TRUE}
bmiClass("a")
```
  
### Extra exercises {#extra}

  - Update survey data with a new variable `heightInch` with height in inch units (use sapply).

```{r}
survey$heightInch <- sapply(survey$height, inch2cm, invert=TRUE)
```
  
  - Update the pulse data with `BMI` variable using the *bmi* function (use apply).

```{r}
d <- pulse[,c("weight","height")]  # intermediate data.frame `d`
pulse$BMI <- apply(d,1,function(x) bmi(x["weight"],x["height"]) )
```

  - Update the pulse data with BMI category variable `bmiClass` using the *bmiClass* function 
  from previous exercise (use sapply).
  
```{r}
pulse$bmiClass <- sapply(pulse$BMI,bmiClass)
```

  
  
