---
title: "Functions and help (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

# Round and signif {#signif}

- Use the `signif` function to give pi in 4 significant digits
```{r}
signif(pi, 4)
```

- (*) What is the value of `round(pi,-1)` and why?
```{r}
round(pi,-1)
```

- (*) For what `x` is `trunc(x)` different from `floor(x)`
```{r}
x=-1.1
trunc(x)
floor(x)
```


## `barplot` options {#help}

Have a look at the `women` example data

- Open the help file for `women` to understand the data.
```{r eval=FALSE}
?women
```

- The `barplot` function can be used for barplots. Make a barplot of `weight`.
```{r}
barplot(women$weight)
```

-	What argument of `barplot` can be used to add names to the bars? HINT: `names.arg`. Add `height` as a names under the bars.
```{r}
barplot(women$weight, names.arg=women$height)
```

- Basic colors are represented in R simply by their names, e.g. `"yellow"`. Note the quotation marks. Make the bars in the plot purple.
```{r}
barplot(women$weight, names.arg=women$height, col="purple")
```

- Add labels to the plot and a title
```{r}
barplot(women$weight, names.arg=women$height, xlab='height', ylab='weight', main='Average weight vs. height') 
```


## Spearman correlation

In the `survey` data:

```{r result="as.is", echo=FALSE, comment="", eval=TRUE}
survey <- catReadTable( "survey.txt", "survey" )
```
```{r eval=FALSE}
survey <- read.delim( "data/survey.txt" )
```

-	The function for correlation in R is `cor`. Calculate correlation between average `span1` and `span2`
```{r}
cor(survey$span1, survey$span1)
```

- Is this a Pearson or Spearman correlation? Calculate the Spearman correlation.
```{r}
cor(survey$span1, survey$span1, method='spearman')
```

# (*) `table` options

`NA` in R stands for missing values

- Open the help file for `table`.
```{r eval=FALSE}
?table
```

- What is the default of the `useNA` argument?
```{r}
"no"
```

- Apply table to the `clap` and `m.i.`columns in `survey` with all three options for `useNA`
```{r}
table(survey$m.i, useNA = 'no')
table(survey$m.i, useNA = 'ifany')
table(survey$m.i, useNA = 'always')
table(survey$clap, useNA = 'no')
table(survey$clap, useNA = 'ifany')
table(survey$clap, useNA = 'always')
```

- Explain the role of the `useNA` argument in your own words.
