---
title: "01-Linear-regression"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```
### Quick Task

Check if the variable `Long.jump` of the `decathlon2` dataset follows a normal distribution using a plot.

```{r}
hist(decathlon2$Long.jump)
qqnorm(decathlon2$Long.jump, pch = 1, frame = FALSE)
qqline(decathlon2$Long.jump, col = "steelblue", lwd = 2)
```

Perform a linear regression the check for association between the variables `Long.jump` and `X100m`. What is the beta value? Is the resul significant?
```{r}
linear.model <- lm(Long.jump ~ X100m, data = decathlon2)
linear.model
summary(linear.model)$coefficients
```

We know that there is a significant difference in mean for the variable `X100m` between both competitions. Add the predictor `Competition` into the model as well as an interaction effect between `Competition` and `X100m` (add X100m:Competition in the formula). Are the results changing?

```{r}
linear.model <- lm(Long.jump ~ X100m + Competition + X100m:Competition, data = decathlon2)
linear.model
summary(linear.model)$coefficients
```