---
title: "Basic data types (cont.)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../../setup/R_course_setup.R" );
```

# Basic data types (cont.) (tasks)

## Numeric vectors {#numeric}

### Primary tasks

- Create a variable `v` as a vector of (46,43,...,19,16,13).
    - Show `v` to check the numbers. 
    - Show the `head` of 4 elements of `v`.
    - Show the `tail` of 3 elements of `v`.
    - Compute the mean of `v`.
    - Calculate the length of `v`.
    
    
```{r}
v <- seq( 46, 13, -3 )
v
head( v, 4 )
tail( v, 3 )
mean(v)
length( v )
```

- Create a variable `v` as a vector of (50, 40, 30, 20, 10, 5, 4, 3, 2, 1, NA). 
    - Use once `seq`, and once `:` (colon). 
    - Show `v` to check the numbers.
```{r}
v <- c( seq( 50, 10, -10), 5:1, NA )
v
```

Compute the mean of `v`.

```{r}
mean(v)
```

- Create a variable `v` as a vector with values 30, 40, 50, 10, 12, 20. 

```{r}
v <- c(30, 40, 50, 10, 12, 20)
```

Compute the mean of `v`.

```{r}
mean(v)
```

Use `sort()` to sort the elements of the vector in ascending order.

```{r}
sort(v)
```

Save the result in another vector `u`. Compute the mean of this vector.

```{r}
u <- sort(v)
mean(u)
```

Check how many elements `v` and `u` have by using `length()`.

```{r}
length(u)
length(v)
```

### Extra tasks

- Sometimes it is useful to create a vector of length _n_ of zeroes. Try the function `numeric( n )` to create a vector of 25 zeroes, then create an empty numeric vector with nothing (so, of zero length). (*)

```{r}
v <- numeric( 25 )
v
v <- numeric()
v
```

- Read the help file of `rep`. Construct a vector 0, 1, 2, 0, 1, 2, 0, 1, 2, 0, 1, 2... with 20 pattern repetitions. (*)
```{r}
rep( 0:2, 20 )
```

- Use `rep`. Construct the first 13 elements of a vector 0, 1, 2, 0, 1, 2, 0, 1, 2, 0, 1, 2... Check the length of the result. (*)
```{r}
v <- rep( 0:2, length.out = 13 )
v
length( v )
```

- Read manual of function `rev`. Use this function and the `v` from the previous task to build a vector of (50,40,30,20,10,5,4,3,2,1,NA,NA,1,2,3,4,5,10,20,30,40,50). (*)
```{r}
v <- 1:5
w <- seq(from = 10, to =50, by = 10)
c( rev(v), rev(w), NA,NA, v, w )
```

## Character vectors {#character}


- Create a character vector `fruits` with 5 random and different fruit names. Check what it contains. 

```{r}
fruits <- c( "apple", "pear", "orange", "lemon", "cherry" )
fruits
```

Check the length of `fruits`.

```{r}
length(fruits)
```

Sort the entries in `fruits` in alphabetical order.

```{r}
sort(fruits)
```

### Extra tasks


- Use the `fruits` vector. Understand `nchar( fruits )`. (*)

```{r}
nchar( fruits )
```

- Build (manually) a vector _idx_ so that `fruits[ idx ]` prints fruit names in alphabetical order. To do this, just make a vector composed of: the index corresponding to the first entry alphabetically; then the index of the second entry alphabetically, and so on. (*)

```{r}
idx <- c( 1, 5, 4, 3, 2 )
```

Use the `idx` vector you created to select elements in `fruits`.

```{r}
fruits[ idx ]
```


- Check what you get when you use `order( fruits )`. (*)

```{r}
order( fruits )
```

Note that this is similar to `idx` you created. So what does the function `order` give?

Use the function `order` to sort the vector `fruits`.

```{r}
fruits[ order(fruits) ]
```

- Sometimes it is useful to create a vector of length _n_ of class `character`. Try the function `character( n )` to create a vector of length 5, then create an empty character vector with nothing (so, with zero length). (*)

```{r}
v <- character( 5 )
v
v <- character()
v
```


- Read the documentation of `rep`. Then understand the code below. (*)

```{r eval=FALSE,echo=TRUE}
c( rep( "male", 2 ), rep( "female", 1 ) )
rep( c( "male", "female" ), c( 2, 1 ) )
```

- Use `rep` to produce an empty character vector with length 5, as you did above with `character`.

## Logical vectors {#logical}

### Primary tasks

- Negation operator: 
    - Practice, what the negation operator produces for these there logical constants: `FALSE`, `TRUE`, `NA`.
    
```{r}
!TRUE
!FALSE
!NA
```

- AND (vector) operator: 
    - Once symmetries are accounted for, there are six possible combinations of two different values (`a`, `b`)  selected from the three logical constants: `FALSE`, `TRUE`, `NA`. Consider all six of them as arguments to the `&` operator. First predict the result of `a & b` and then calculate it.
    
```{r}
F & F
F & T
F & NA
T & T
T & NA
NA & NA
```

- Make a vector `v <- 19:10`. Build a logical vector `w` which states which positions of `v` are smaller than 12 or at least 17. Print both `v` and `w` to check by eye whether the result is correct.

```{r}
v <- 19:10
v
w <- (v < 12) | (v >= 17)
w

```

- Logical vectors can also be transformed into numeric. Do this for the vector `w` created above.

```{r}
as.numeric(w)
```

What happens to the entries in `w` equal to TRUE? And those equal to `FALSE`? 

- Count how many positions in `w` are equal to `TRUE`. For this, use the `sum` function.

```{r}
w
sum( w )
```



### Extra tasks

- OR (vector) operator: (*)
    - Reproduce the "AND operator" exercise for the `|` (or) operator.

- Elements on `which` positions (indices) are `TRUE`: (*)
    - Study the logical vector `w` from the earlier task. 
    - First, answer yourself the question: which positions in `w` are `TRUE`. 
    - Was your answer the same as the result of `which( w )`? 
    - Can you explain `which( LETTERS == "R" )`?
```{r}
which( w )
which( LETTERS == "R" )
```

- Even numbers: (*)
    - Even numbers are divisible by two with rest zero.
    - Rest of division can be calculated with the `%%` operator. 
    - Take the vector `v <- 19:10` and construct the corresponding logical vector `w` stating whether elements of `v` are divisible by two.
    
```{r}
v <- 19:10
v
v %% 2
v %% 2 == 0
w <- v %% 2 == 0
w
```

- AND, OR (single/first value) operators `&&`, `||`: (*)
    - They work on single values (instead of complete vectors) and are frequently used in programming.
    - Check the manual ``?`&&` ``. Try with some logical values similarly to `&` and `|`.

- AND, OR (single/first value) functions `all`, `any`:
    - Programs are more readable, when instead of `a && b && c && d` the following code is written: `all( a, b, c, d )`. 
    - Read the manual for `all`, `any`. (*)


## Type conversions {#conversions}

### Primary tasks

- Take the following character vector: `v <- c( "0", "1", "11", "121", "1221", "one" )`. Check the result of `as.numeric( v )`.

```{r}
v <- c( "0", "1", "11", "121", "1221", "one" )
v
as.numeric( v )
```

- Produce texts (character vector) out of numbers `101:110`.
```{r}
as.character( 101:110 )
```

- Convert `c( -1:1, NA )` to a logical vector with `as.logical`. Explain the result.
```{r}
# NA stays NA, zero becomes FALSE, other values become TRUE
as.logical( c( -1:1, NA ) )
```

### Extra tasks

- Which of elements of `c( "TRUE", "True", "true", "T", "FALSE", "False", "false", "F" )` are converted to valid logical values with `as.logical`? (*)
```{r}
as.logical( c( "TRUE", "True", "true", "T" ) )
as.logical( c( "FALSE", "False", "false", "F" ) )
as.logical( c( "f", "t", "F", "T" ) )
```

- Sometimes it is useful to create a vector of given number _n_ of `FALSE` values. Try the function `logical( n )` to create a vector of 5 `FALSE`s, then create an empty logical vector. (*)
```{r}
v <- logical( 5 )
v
v <- logical()
v
```
