---
title: "Selecting/indexing/subsetting (vector elements, data frame rows/columns, ...) (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../../setup/R_course_setup.R" );
```

# Selecting/indexing/subsetting (vectors) (tasks)

## Square brackets operator {#brackets}

Define a vector of the sequence of numbers between 101 and 110.

```{r echo=TRUE,eval=TRUE}
v <- 101:110
```

### Primary tasks

- Get the seventh element of the vector. 

```{r}
v[ 7 ]

```

- Get the number of entries in the vector (the same as the length of the vector).

```{r}
length(v)
```

- Get the last element of the vector.

```{r}
v[ length( v ) ]
```

- Check what happens when you try to get an element with a too large index - 20, say.

```{r}
v[ 20 ]
```

- Check what happens when you use a negative index - say -7. What can negative indices be used for?

```{r}
v[ -7 ]
```

- Set the 5th element of `v` to zero. Check `v` before and after this change.
```{r}
v
v[ 5 ] <- 0
v
```

- Create a vector of indices `idx` with the entries 3, 7 and 1. Get the entries of `v` for positions indicated by `idx` - so the third, seventh and the first.

```{r}
idx <- c( 3, 7, 1 )
v[ idx ]
```

### Extra tasks

- Name the elements of `v` with subsequent small letters of alphabet. Use `1:length( v )` to specify the range of the `letters` vector. (*)

```{r}
names( v ) <- letters[ 1:length( v ) ]
v
```

- Get the values of `v` corresponding to positions named `e` and `h`. (*)
```{r}
v[ c( 'e', 'h' ) ]
```

- In one operation set the 3rd and 5th elements of `v` to 13 and 15. Check the result. (*)

```{r}
v
v[ c( 3, 5 ) ] <- c( 13, 15 )
v
```


## Through utility functions {#utility}

### Result is (part of) the original vector

#### Primary tasks

- Sample a vector randomly: `sample`.
    - Define `v <- 101:110`. Try several times `sample( v )`. 
    - Then try several times `sample( v, 3 )`, `sample( v, 15 )` and `sample( v, 15, replace = TRUE )`.
 
```{r error=TRUE}
v <- 101:110
sample( v )
sample( v )
sample( v, 3 )
sample( v, 3 )
sample( v, 15 )
sample( v, 15, replace = TRUE )
sample( v, 15, replace = TRUE )
```

   - Using one result of sampling yielding 15 values, select the first 3. 

```{r}
vs <- sample(v, 15, replace = TRUE)
vs[c(1, 2, 3)] # select the first 3 elements
vs[1:3] # Same result as before
```
   

#### Extra tasks


- Reverse a vector: `rev`. (*)
    - Define `v <- 101:110`. 
    - Compare `v` to `rev( v )`.
```{r}
v <- 101:110
v
rev( v )
```

### Result is a logical vector

#### Primary tasks

- Which element(s) are available/missing: `is.na`.
    - Create a random vector `v` with `NA` elements: `v <- sample( c( 0, 1, NA ), 10, replace = TRUE )`. 
    - Understand `is.na( v )` and `v[ !is.na( v ) ]`.
```{r}
v <- sample( c( 0, 1, NA ), 10, replace = TRUE )
v
is.na( v )
v[ !is.na( v ) ]
```

#### Extra tasks

- Whether elements of one vector are present in another vector: `%in%`. (*) 
    - Let's check whether elements of `v <- c( "1", "B", "I", "O", "b" )` are among `LETTERS` with `v %in% LETTERS`. 
    - How to interpret the result? 
    - How to find how many of elements in `v` occur in `LETTERS`?
    
```{r}
v <- c( "1", "B", "I", "O", "b" )
v
v %in% LETTERS
sum( v %in% LETTERS )
```

- Whether elements of a vector are duplicated (have occurred earlier already): `duplicated`. (*)
    - Let's generate a vector `v` with repeated elements `v <- sample( 1:4, 10, replace = TRUE )`. 
    - Understand `duplicated( v )`.
    
```{r}
v <- sample( 1:4, 10, replace = TRUE )
v
duplicated( v )
```

### Result is a vector of indices

#### Primary tasks

To find the index of the largest/smallest entry in a vector, you can use: `which.max`, `which.min`.
    
- Create a vector `v` by sampling 10 elements from the sequence `1:1000000`. Show `v`.
    
    
```{r}
v <- sample( 1:1000000, 10 )
v
```

- Use `which.max( v )` to find which entry of `v` contains its largest element. 

```{r}
which.max( v )
```

- Select the largest element of `v`.

```{r}
v[ which.max( v ) ]
```

- Check that this is indeed the largest value in `v` by using `max`.

```{r}
max(v)
```

#### Extra tasks

- Indices where a logical vector is `TRUE`: `which`. (*)
    - Let's create a random logical vector `v <- sample( c( T, F ), 10, replace = TRUE )`. 
    - Use `which` to find positions with `TRUE` values. 
    - What is `v[ which( v ) ]`?
```{r}
v <- sample( c( T, F ), 10, replace = TRUE )
v
which( v )
v[ which( v ) ]
```

- Indices at which elements of one vector `match` another vector: `match`. (*)
    - Let's check where elements of `v <- c( "1", "B", "I", "O", "b" )` are among `LETTERS` with `match( v, LETTERS )`.
    - How to interpret the result? 
    - How to find how many of elements in `v` do not occur in `LETTERS`?
```{r}
v <- c( "1", "B", "I", "O", "b" )
idx <- match( v, LETTERS )
idx
LETTERS[ idx ]
sum( is.na( idx ) )
```

- Indices ordered so, that the elements of a vector would get ordered: `order`. (*)
    - Let's generate a random vector `v <- sample( LETTERS, 10 )`.
    - Understand `order( v )` and `v[ order( v ) ]`.
    - Check whether `v[ order( v ) ] == sort( v )` or even better whether `all( v[ order( v ) ] == sort( v ) )` is TRUE.
```{r}
v <- sample( LETTERS, 10 )
v
order( v )
v[ order( v ) ]
v[ order( v ) ] == sort( v )
all( v[ order( v ) ] == sort( v ) )
```

### Other functions

- Sorting a vector
    - Define `v` as a  sample of small letters of size 5. Check its content. 

```{r}
v <- sample( letters, 5)
v
```

- Sort `v` using `sort()`. Save the result in a new vector.

```{r}
sort( v )
u <- sort(v)
```

- Now computer `order(v)`.

```{r}
order(v)
```

To understand what the function `order()` does, check what happens when you use it to select entries in `v`:

```{r}
v[ order(v) ]
```

Compare the result to `u`.

```{r}
u
```

- Obtaining all unique elements of a vector
  - Generate again a vector `v` with the sequence of numbers between 101 and 110.
    
```{r}
v <- 101:110
```

  - Sample 15 elements of this vector with replacement, as done in a previous exercise.

```{r}  
vs <- sample( 101:110, 15, replace = TRUE )
```

  - Now use `unique()` to obtain the set of unique entries in the sampled vector.
    
```{r}
unique( vs )
```

  - Use a function on this result to find out how many unique values the sampled vector has.
    
```{r}
length(unique( vs ))
```


- Obtaining the logical vector indicating duplicated entries (those which have occurred earlier)
  - Use `duplicated()` to obtain the logical vector indicating which entries are duplicated in `v`. Check if the result makes sense by comparing it to `v` itself

```{r}
duplicated(v)
v
```

  - Use the logical vector created to select entries of the vector `v` which are duplicated
    
```{r}
v[ duplicated(v) ]
```

  - Compute the number of duplicated entries in `v`.

There are multiple ways to find out how many entries in `v` are duplicated. Some of them are:

```{r}
sum( duplicated( v ) )
length( v[ duplicated(v) ] )
```

#### Extra tasks

- Create a vector with ranks of elements of another vector: `rank`.
    - Let's generate a random vector `v <- sample( LETTERS, 10 )`.
    - Understand the difference of `order( v )` and `rank( v )`.
```{r}
v <- sample( LETTERS, 10 )
v
order( v )
rank( v )
```

*Explanation*: We already know that `order` returns the indices that, when used on the original vector, sorts its elements. The function `rank` gives you the relative position of each entry, from the smallest to the largerst by default. For example, define

```{r}
myv <- c("A", "C", "E", "B", "F")
myv
```

It is easy to see that the result of `order` should be the vector: `c(1, 4, 2, 3, 5)`, since the first element in alphabetical order `A` is at position one, then the second element in alphabetical order `B` is in position 4, etc. Indeed, check that this is also the result of `order`:

```{r}
order(myv)
```

The `rank` gives you the relative position of each entry, also in alphabetical order. The first entry in `myv` is also the first element in alphabetical order, so the first element in `rank` is 1. The second entry in `myv` is the fourth element in alphabetical order, so the second element in `rank` is 4.  Then the third entry in `myv` is the second element in alphabetical order, with rank 2. The next entry in `myv` is the third element, and the last entry in `myv` is the fifth element. This means that `rank(myv)` returns `c(1, 4, 2, 3, 5)`. Check this:

```{r}
rank(myv)
```
