---
title: "Data frames (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../../setup/R_course_setup.R" );
```

# Data frames (tasks)

## Tasks

### Creating a data frame

#### Primary tasks

- Create a `data.frame` from given vectors

Create the following vectors: `firstname` containing "George", "Leila" and "Alice";  `age` containing 34, 27 and 41; and `height` containing 173, 177 and 162. 

```{r}
firstname <- c("George", "Leila", "Alice")
age <- c(34, 27, 41)
height <- c(173, 177, 162)
```

Create a `data.frame` containing these variables.

```{r}
mydata <- data.frame(firstname, age, height)
```

Check the structure of the `data.frame` created.

```{r}
str(mydata)
```

Compute the mean height of the individuals in the data.

```{r}
mean(mydata$height)
```

Compute the mean age of the individuals in the data.

```{r}
mean(mydata[, "age"])
```

Check what happens if you try to compute the mean by selecting the variable `age` via square brackets, without the comma:

```{r error = TRUE}
mean(mydata[, "age"])
```

### Reading data from files with different formats

#### Primary tasks

- Load the `pulse.csv` file into the `d` object. Display the structure of `d` and the first few rows.

```{r eval=FALSE}
d <- read.csv( "data/pulse.csv" )
```

```{r result="as.is", echo=FALSE, comment="",eval=TRUE}
# Do not run this code chunk during the course
d <- catReadCsv( "pulse.csv", "d" )
```

```{r}
str( d )
head( d )
```

Tabulate `gender` and `exercise`.

```{r}
table(d$gender)
table(d$exercise)
```

- Select a small part of the data to work with

Select the first 10 rows from table `d`. 
Save the data for the selected rows as a new object called `dnew`.

```{r}
dnew <- d[1:10, ]
```

Compute the mean height in this selected data.

```{r}
mean(dnew$height)
```

Use the function `summary()` on the `height` variable.

```{r}
summary(dnew$height)
```

Make a new data.frame `d2` with the same selected rows and only columns `height`, `weight` and `age`.

```{r}
d2 <- dnew[, c("height", "weight", "age")]
```

Use the function `summary()` on this smaller dataset. What do you obtain?

```{r}
summary(d2)
```


- Sorting the rows of a data.frame by one or more variables

Use the dataset `d2` you just created. Use `sort()` to sort the entries of `age`.

```{r}
sort(d2$age)
```

Now use `order()` to again sort the elements of `age`. Hint: you have already done this in `02_selecting.tasks`.

```{r}
d2$age[ order(d2$age) ]
```

Use now the same trick to sort the rows in `d2`, including all columns, according to `age`.

```{r}
d2[ order(d2$age), ]
```

Now sort the rows in `d2` first by `age` and then by `weight`. This means: individuals of the same `age` are sorted by their `weight`.

```{r}
d2[ order(d2$age, d2$weight), ]
```

Sort the rows in `dnew`, the data.frame from the previous exercise which contains the same individuals as `d2` but with more columns, according to both `age` and `weight`.

```{r}
dnew[ order(d2$age, d2$weight), ]
```

Now display the sorted rows only for the columns `name` and `year`.

```{r}
dnew[ order(d2$age, d2$weight), c("name", "year")]
```


- Selecting rows by a logical condition.
    
You have been asked to select individuals from the `d` data corresponding to men of weight in the range 60-80 kg. One way of doing this is as follows. 

1. Create a logical vector `isMan` stating whether a row in the data corresponds to men. Hint: check which values in `gender` correspond to men.

```{r}
table(d$gender)
isMan <- d$gender == "male"
```

Tabulate the logical vector to check you have constructed it correctly.

```{r}
table(isMan)
```

2. Create a logical vector `isInWeightRange` stating whether a row corresponds to a person with weight between 60 and 80

```{r}
isInWeightRange <- (d$weight >= 60) & (d$weight <= 80)
```

Tabulate the resulting vector.

```{r}
table(isInWeightRange)
```

3. Select rows of the object `d` corresponding to the combined selections

```{r}
d[ isMan & isInWeightRange, ]
```



#### Extra tasks

- Enforce a data frame type.
    - Observe below, that `d[ ,cols ]` might sometimes return a vector, and sometimes return a data frame.
    - Adding `drop = FALSE` argument enforces returning a data frame object.
```{r echo=TRUE}
cols <- c( "weight", "height" )
class( d[ , cols ] )
cols <- c( "weight" )
class( d[ , cols ] )
cols <- c( "weight" )
class( d[ , cols, drop = FALSE ] )
```

