---
title: "Matrices (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE,eval=TRUE}
source( "../../setup/R_course_setup.R" );
```

# Matrices (tasks)

## Tasks

### Matrix creating

Create a matrix with 5 rows and 5 columns, with entries formed by the sequence of integers from 1 to 25. Fill in entries by row: first entries on row 1, then entries on row 2, and so on. Save the matrix as an object. Show the matrix created.

```{r}
mat <- matrix(1:25, nrow = 5, ncol = 5, byrow = TRUE)
mat 
```

Assign the following row names to this matrix: "Row 1", "Row 2", and so on until "Row 5".

```{r}
rownames(mat) <- c("Row 1", "Row 2", "Row 3", "Row 4")
```

Compute the sum of columns of the matrix, per column.

```{r}
colSums(mat)
```

Compute the row means of the matrix, per row.

```{r}
rowMeans(mat)
```

You have been told that the element on the second row and fifth column was actually not observed. Make this entry `NA`.

```{r}
mat[2, 5] <- NA
```

Now recompute the column sums. Check the help file of the function used, to make sure that `NA`s are ignored when computing the sums.

```{r}
# ?colSums
colSums(mat, na.rm = TRUE)
```

Use the function `summary()` to compute various statistics of the matrix columns. 

```{r}
# ?summary
summary(mat)
```

### Matrix of correlation coefficients {#corMatrix}

#### Extra tasks

The goal of this exercise is to calculate a matrix of column-column Pearson correlation coefficients for a number of data frame columns. The `cor` function does not accept a `data.frame` argument and it works only with numerical columns. Therefore, first the numerical columns of the data frame need to be extracted and converted to a matrix.

- Read the pulse data frame to `pulse` variable; check its class.
```{r}
pulse <- read.table( "data/pulse.txt" )
```
```{r result="as.is", echo=FALSE, comment=""}
catReadTable( "pulse.txt", "pulse" )
```

```{r}
class( pulse )
```

- Calculate correlation of columns `height` and `weight`.
```{r}
cor( pulse$height, pulse$weight )
```

- Understand the error of `cor( pulse )`.
```{r error=TRUE}
cor( pulse )
```

- Select only numerical columns of `pulse`. Then use `as.matrix` on the selected columns. Store the result in `m`. Confirm that `m` is a matrix. Check `m` dimensions and `str( m )`.
```{r}
m <- as.matrix( pulse[ c( "height", "weight", "age", "pulse1", "pulse2", "year" ) ] )
class( m )
dim( m )
str( m )
```

Note that it does not make sense to compute the correlation with `year`, although this is technically allowed.

- Try `cor( m )` and understand the result. Read the `cor` documentation for further ideas related to `NA`. Assign the final result to `corM`.
```{r}
cor( m )
cor( m, use = 'p' )
corM <- cor( m, use = 'p' )
```

- Read documentation for `round` and round the `corM` matrix to 2 most significant digits. (*)
```{r}
round( corM, 2 )
```
