---
title: "02-log-rank-test"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```


### Quick tasks

1. Consider again the leukemia data in dataset `leuk`. We will now compare survival probabilities between the groups given by `ag`, representing the test result for the presence of Auer rods and/or significant granulation of leukaemic cells.

Plot the Kaplan-Meier survival curves separately per group defined by the AG test. Compute the log-rank test to compare the two survival distributions between the two groups.

```{r}
# Create the Surv object
leuk.surv <- Surv(leuk$time) # no event variable as all patients experience the event (death)
# Define the comparison to be made
leuk.fit <- survfit(leuk.surv ~ leuk$ag)
# Make a plot if the survival distributions for the two groups
mycols <- c("blue", "purple")
plot(leuk.fit, col = mycols)
legend("topright", legend = levels(leuk$ag), fill = mycols)
# Test whether the difference is statistically significant using the log-rank test
survdiff(leuk.surv ~ leuk$ag)
sdif <- survdiff(leuk.surv ~ leuk$ag)$chisq
lr.p <- pchisq(sdif, df = 1, lower.tail = FALSE)

# Make a plot if the survival distributions for the two groups
mycols <- c("blue", "purple")
plot(leuk.fit, col = mycols, main = paste("log-rank p=", round(lr.p, 3)))
legend("topright", legend = levels(leuk$ag), fill = mycols)

```