---
title: "03-Survival-analysis-regression-models"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```


### Quick tasks


1. Make Kaplan-Meier curves to the Melanoma data using the same `Surv` object defined above, i.e. with both censoring categories combined into a single one. Do this both for the entire data together, as well as for the observations according to gender. Compute the log-rank test to check whether there is evidence that the two survival probability curves are different.



```{r}
survm <- Surv(Melanoma$time, Melanoma$status == 1)
```


```{r}
plot(survfit(survm ~ 1))
```

```{r}
plot(survfit(survm ~ Melanoma$sex))
```

```{r}
survdiff(survm ~ Melanoma$sex)
```

2. Let us consider again the `leuk` example with overall survival data of leukemia cases, of the package `MASS`. 

```{r}
library(MASS)
library(survival)
str(leuk)
```

We define again the response as a function of `time` only, since all patients experience an event (death). Subsequently, we fit a survival regression of the survival response as a function of `ag` and of white blood cell count `wbc`. We will use Cox regression for this.

First, let us examine the nature of the relationship between `time` and the covariates `ag` and `wbc`. For this, we make a graph. We use `time` on the log-scale, as this is equivalent to what happens in the regression model:

```{r, fig.width = 10, fig.height = 5}
par(mfrow = c(1, 2))
plot(log(time) ~ ag + wbc, data = leuk)
```

The covariate `wbc` has many small values and only a few large ones. This is characteristic of a variable on an exponential scale. So we take its logarithm and produce the plot again:

```{r}
plot(log(time) ~ log(wbc), data = leuk)
```

This relationship seems to be closer to linear. So, we will include `wbc` in the model with a log-transformation.

A Cox regression is fitted via the function `coxph` which, as regression functions `lm` and `glm`, expect a formula as input. The formula relates the `Surv` response to the covariates, in this case `ag` and `log(wbc)`:

```{r}
leuk.cox <- coxph(Surv(time) ~ ag + log(wbc), data = leuk)
summary(leuk.cox)
```

3. Consider again the Melanoma data, and now define `sex` as 1 for females and 0 for males. Fit the Cox regression model again and compare results with those when `sex` was used with the original coding. 

```{r}
sex1 <- 1 - Melanoma$sex
mel.cox1 <- coxph(survm ~ sex1)
summary(mel.cox1)
# With the original coding we had:
mel.cox
```
