---
title: "Data frames (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup,include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

# Data frames (tasks)

## Tasks

### Reading a different format

#### Primary tasks

- Load the `pulse.csv` file into the `d` variable. Display the structure of `d` and the first few rows.
```{r eval=FALSE}
d <- read.csv( "data/pulse.csv" );
```
```{r result="as.is", echo=FALSE, comment="",eval=TRUE}
d <- catReadCsv( "pulse.csv", "d" )
```
```{r}
str( d )
head( d )
```

- Random sampling of rows:
    - Sample randomly 10 rows from table `d`. 
    - First create a variable `rows` which will contain the row numbers.
    - Show selected rows.
```{r}
rows <- sample( 1:nrow( d ), 10 )
d[ rows, ]
```

- Sorting (ordering) a table (by multiple keys):
    - Use `View( d )` to visualize the `d` data frame.
    - Understand: `rows <- order( d$age )`. Use `View` to visualize `d[ rows, ]`.
    - Understand: `rows <- order( d$age, d$weight )`. Use `View` to visualize `d[ rows, ]`.
    - Finally: display only columns `name`, `age` and `weight` of sorted `d`.
    - Note: for clarity, the solutions use only random 10 rows of `d`.
```{r}
set.seed( 1000 ) # for reproducible generation of random numbers
rows <- sample( 1:nrow( d ), 10 )
dd <- d[ rows, ]
dd
rows <- order( dd$age )
dd[ rows, ]
rows <- order( dd$age, dd$weight )
dd[ rows, ]
dd[ rows, c( "name", "age", "weight" ) ]
```

- Selecting rows by a logical condition.
    - Goal: select rows with males of weight range 60-80
    - Build a logical vector `isMale` stating whether a row describes a male
    - Build a logical vector `isInWeightRange` stating whether a row describes a person with weight between 60 and 80
    - Select rows which are `isMale & isInWeightRange`
    - Note: for clarity, the solutions use only random 10 rows of `d`.
```{r}
set.seed( 1000 ) # for reproducible generation of random numbers
rows <- sample( 1:nrow( d ), 10 )
dd <- d[ rows, ]
isMale <- dd$gender == "male"
isMale
isInWeightRange <- dd$weight >= 60 & dd$weight <= 80
isInWeightRange
isMale & isInWeightRange
dd[ isMale & isInWeightRange, ]
```

#### Extra tasks

- Enforce a data frame type.
    - Observe below, that `d[ ,cols ]` might sometimes return a vector, and sometimes return a data frame.
    - Adding `drop = FALSE` argument enforces returning a data frame object.
```{r echo=TRUE}
cols <- c( "weight", "height" )
class( d[ , cols ] )
cols <- c( "weight" )
class( d[ , cols ] )
cols <- c( "weight" )
class( d[ , cols, drop = FALSE ] )
```

