---
title: "Matrices (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE,eval=TRUE}
source( "../R_course_setup.R" );
```

# Matrices (tasks)

## Tasks

### Matrix of correlation coefficients {#corMatrix}

#### Extra tasks

The goal of this exercise is to calculate a matrix of column-column Pearson correlation coefficients for a number of data frame columns. The `cor` function does not accept a `data.frame` argument and it works only with numerical columns. Therefore, first the numerical columns of the data frame need to be extracted and converted to a matrix.

- Read the pulse data frame to `pulse` variable; check its class.
```{r}
pulse <- read.table( "../data/pulse.txt" )
```
```{r result="as.is", echo=FALSE, comment=""}
catReadTable( "pulse.txt", "pulse" )
```

```{r}
class( pulse )
```

- Calculate correlation of columns `height` and `weight`.
```{r}
cor( pulse$height, pulse$weight )
```

- Understand the error of `cor( pulse )`.
```{r error=TRUE}
cor( pulse )
```

- Select only numerical columns of `pulse`. Then use `as.matrix` on the selected columns. Store the result in `m`. Confirm that `m` is a matrix. Check `m` dimensions and `str( m )`.
```{r}
m <- as.matrix( pulse[ c( "height", "weight", "age", "pulse1", "pulse2" ) ] )
class( m )
dim( m )
str( m )
```

- Try `cor( m )` and understand the result. Read the `cor` documentation for further ideas related to `NA`. Assign the final result to `corM`.
```{r}
cor( m )
cor( m, use = 'p' )
corM <- cor( m, use = 'p' )
```

- Read documentation for `round` and round the `corM` matrix to 2 most significant digits. (*)
```{r}
round( corM, 2 )
```
