---
title: "Basics (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../../setup/R_course_setup.R" )
```

# First calculations {#firstCalculations}

- Calculate 25 plus the square of 12
```{r}
25 + 12^2
```

- What is the square root of the number you have just calculated?
```{r}
sqrt(25 + 12^2)
```



# First variables {#firstVariables}

- Make a variable that contains your height in m.
```{r}
height_m <- 1.78
```

- Calculate your height in cm.
```{r}
height_m*100
```

- Store the result of the previous calculation as a variable
```{r}
height_cm <- height_m/100
```


# Making a vector {#exercisesBasics}

Four patients had dietary intakes of 2314, 2178, 1922, 2004 kcalories.

- Make a vector of these four values.
```{r}
diet_kcal <- c(2314, 2178, 1922, 2004)
```
- Convert these values into in KJ using 1 kca = 4.184 kJ
```{r}
diet_kJ <- diet_kcal * 4.184
```
- (*) Sort the resulting values using the function `sort`.
```{r}
sort(diet_kJ)
```


# Regular sequences

-	Make a vector of all numbers between 1 and 100. Hint: use `:`.
```{r}
x <- 1:100
```

- Calculate 1+2+3+...+100. Hint: use `sum`.
```{r}
sum(x)
```


# Vectorized calulation

- Degrees Celsius (Cel) can be calculated from degrees Fahrenheit (Fah) as Cel = 5/9 * (Fah - 32). How much is 60 degrees Fahrenheit in Celsius?
```{r}
5/9 * (60 - 32)
```

-	Calculate a vector that contains degrees Celsius for 0,1,...,100 degrees Fahrenheit.
```{r}
Fah <- 1:100
Cel <- 5/9 * (Fah - 32)
```

- From your calculation in the previous exercise, select 10 Fahrenheit in Celcius.
```{r}
Cel[10]
```


# (*) Removing variables

- R knows the value of `pi`. Type `pi` to check
```{r}
pi
```

- Overwrite pi with a variable of your own 
```{r}
pi <- 3
pi
```

- Remove and check that the old pi is available again 
```{r}
rm(pi)
pi
```


# (*) Integer division and vectors

- Use `%/%` to make a vector of length 16 with values 1,1,1,1,2,2,2,2,3,...
```{r}
1:16 %/% 4
```

- Use `%%` to make a vector of length 16 with values 1,2,3,4,1,2,3,4,1,...
```{r}
1:16 %% 4 + 1
```




