---
title: "Data frames (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../../setup/R_course_setup.R" )
```

# Explore {#explore}

For the quick exercises we will mostly work with the `survey` data. Let's load them into R.



- Download the data in `survey.txt` and load them into R. Use the script you made!
```{r eval=1:2, echo=3}
survey <- read.delim("../../data/survey.txt")
pulse <- read.delim("../../data/pulse.txt")
```

- How many rows and columns does this data set have?
```{r}
dim(survey)
```

- Which variables are in the data?
```{r}
names(survey)
```



# Data frame columns {#columns}

In the `survey` data 

- Calculate the difference between spans of the the writing and non-writing hands for each student
```{r}
diff <- survey$span1 - survey$span2

```

- Add the result of your calculation as a new column to your data set
```{r}
survey$diff <- diff
```

- What is the minimum age in the data set? Hint: use `min`.
```{r}
min(survey$age)
```

In the `pulse` data

- BMI is weight in kg divided by the square of height in meters. Add a column BMI for each student. HINT: in the pulse data, the `height` variable is given in cm.
```{r}
pulse$BMI <-pulse$weight/(pulse$height/100)^2
```

- (*) Remove the BMI column you have just added to the pulse data.
```{r}
pulse$BMI <- NULL
```


# Exploring a new data.frame {#newFrame}

The `swiss` data set has demographic data on swiss provinces/regions around 1888. It is an example data.frame that is available in R. Just type `swiss` at the prompt to see it.

-	Use the functions View, head and tail to get an overview of the data set
```{r}
head(swiss)
tail(swiss)
```

- Have a look at the help file. What is this data set about?
```{r eval=FALSE}
?swiss
```

- How many provinces are in this data set?
```{r}
nrow(swiss)
```

-	What is the largest percentage of catholics in any of these Swiss provinces? Hint: use `max`
```{r}
max(swiss$Catholic)
min(swiss$Catholic)
max.cath <- max(swiss$Catholic)
```

- Find the Fertility of `Lavaux`.
```{r}
swiss["Vevey", "Examination"]

swiss["Lavaux", "Fertility"]

```

- Make a new column in the swiss data that contains the percentage of protestants (100 minus the percentage of catholics)
```{r}
swiss$Protestant <- 100 - swiss$Catholic
```



