---
title: "Basic data types (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../../setup/R_course_setup.R" );
```

```{r echo=FALSE}
pulse <- read.delim("../../data/pulse.txt")
survey <- read.delim("../../data/survey.txt")
```

# Numeric data {#numeric}

In the `survey` data

- Find the standard deviation of `span1`
```{r}
sd(survey$span1)
```

- Find the age range
```{r}
range(survey$age)
```

- plot `span1` (y-axis) against `span2` (x-axis)
```{r}
plot(survey$span2, survey$span1)
```

- (*) plot the difference between `span1` and `span2` (y-axis) against their average (x-axis)
```{r}
survey$diff.span <- survey$span1 - survey$span2
survey$mean.span <- (survey$span1 + survey$span2)/2
plot(survey$mean.span, survey$diff.span)
```

# Factors {#factors}

In the `survey` data

- check the class of the `fold` variable

```{r}
class(survey$fold)
```

- Transform the `fold` variable into a factor. How many and what levels does it have?
```{r}
survey$nfactor <- as.factor(survey$factor)
levels(factor(survey$nfold))
```

- Make a `table` of the `fold` variable (columns) versus the `clap` variable (rows)
```{r}
table(survey$clap, survey$fold)
```

- Convert the variables `hand`, `exercise` and `smoke`  to `factor`
```{r}
survey$hand <- as.factor(survey$hand)
survey$exercise <- as.factor(survey$exercise)
survey$smokes <- as.factor(survey$smokes)
```

- Do the same for variables `gender`, `smokes`, `alcohol`, exercise and `ran` from the `pulse` data.

- Add these steps to your standard script for reading in the data! 


# Descriptive statistics of a data set

We make use of the `cats` data set from the `MASS` package. 

-	Load the `MASS` package by writing `library(MASS)`. This makes the `cats` data available to you. What variables of what types are in `cats`?
```{r}
library(MASS)
str(cats)
```

- How many male and female cats were in this experiment?
```{r}
table(cats$Sex)
```

- Make a histogram of heart weights
```{r}
hist(cats$Hwt)
```

- Find mean and median heart weight
```{r}
mean(cats$Hwt)
median(cats$Hwt)

```

-	Plot heart weight against body weight
```{r}
with(cats, plot(Bwt, Hwt))
```

-	(*) `plot` when applied to a factor and a numeric variable makes boxplots. Make boxplots of body weight against gender
```{r}
plot(cats$Sex, cats$Bwt)
```

