---
title: "03-Testing-independence-in-nx2-tables"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```

### Quick tasks

1. Use the chi-square test to decide if there is evidence of association between learning speed and age in the `quine` data. Extract the p-value of the test, and add the p-value to the barplot of the data.

```{r}
mych.test <- chisq.test(quine$Lrn, quine$Age)
```


```{r}
plot(quine$Lrn ~ quine$Age, main = paste("Chi-square p=", round(mych.test$p.value, 2) ) )
                                         
```


2. Now consider two other datasets involving variables `X` and `Y`, with the cross-tabulations of the variables given below - first for study 1:


```{r}
table(x1, y1)
```

Now for study 2:

```{r}
table(x2, y2)
```

Use a test to investigate whether or not the data supports the null hypothesis of independence between `X` and `Y`.

```{r}
myt1 <- table(x1, y1)
chisq.test(myt1)

myt2 <- table(x2, y2)
chisq.test(myt2)
```

