---
title: "08-power-analysis-poisson-regression"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```


### Quick tasks

Compute the power for a series of multiplying factors for the column sums of the learning speed and sex table of the `quine` data, considering now the problem of comparing proportions of average learners per group. Plot the estimated powers and determine the minimum sample size corresponding to 0.80 power.

```{r}
sr <- seq(from = 1, to = 2.5, by = 0.1)
powerv <- NULL
for(xi in 1:length(sr)) powerv <- c(powerv, 
                                    power.fisher.test(p1  = 40/80, p2 = 23/66, 
                                                      n1 = floor(80*sr[xi]), n2 = floor(66*sr[xi])))

```

The graph of the estimated power, as a function of the multiplying factor:

```{r}
plot(sr, powerv, pch = 20, col = "blue", main = "Power Fisher exact test", 
     xlab = "factor multiplying both marginal counts", ylab = "estimated power")
```

So, a power of at least 0.80 is achieved for the last few multiplying factors, corresponding to minimum sample sizes required of:

```{r}
min(sr[ powerv >= 0.8 ]*80)
```

for group 1, and 

```{r}
min(sr[ powerv >= 0.8 ]*66)
```

for group 2.