---
title: "Basic statistical tests (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE, eval=F}
source( "../R_course_setup.R" );
```

```{r echo=FALSE}
pulse <- read.delim("../data/pulse.txt")
survey <- read.delim("../data/survey.txt")
```

# wilcox.test {#S3}

- Perform a (paired) Wilcoxon signed rank test comparing `span1` and `span2` in the survey data

```{r}
wilcox.test(survey$span1, survey$span2, paired=TRUE)
```

- Store the p-value and the test statistic as two separate variables

```{r}
res <- wilcox.test(survey$span1, survey$span2, paired=TRUE)
names(res)
p <- res$p.value
V <- res$statistic
```

- Apply now the Wilcoxon test ignoring the paired nature of the data and save the p-value generated as a variable. Then print on the console the two p-values computed (using the pairing and then not using it) on two subsequent lines.


```{r}
res.np <- wilcox.test(survey$span1, survey$span2, paired=FALSE)
names(res.np)
p.np <- res.np$p.value
p
p.np
```

# Unpaired t.test

- Study the `t.test` help file. How should we supply the input if we want to do a t test to compare `span1` between males and females?

```{r eval=FALSE}
?t.test
```

- Make two separate vectors of `span1` values for males and females.
```{r}
span1.male <- survey$span1[survey$gender == 'male']
span1.female <- survey$span1[survey$gender == 'female']
```

- Perform the unpaired t test that compares span1 between males and females.

```{r}
t.test(span1.male, span1.female)
```

- Extract the confidence interval for the difference in spans from the resulting object.
```{r}
res <- t.test(span1.male, span1.female)
names(res)
res$conf.int
```

- Round the p-value to 3 significant digits. Hint: use `round`
```{r}
round(res$p.value, 3)
```


# Tests for count tables 

- In the survey data, make a table of the `fold` against the `clap` variables. Hint: use the function `table`.
```{r}
table(survey$fold, survey$clap)
```

- Now use a chi-square test to study the association between `fold` and `clap`. For this, check the help file for the function `chisq.test`.  

```{r}
chisq.test(survey$fold, survey$clap)
```

- Check the class and the names of this object. Save the computed p-value as a separate variable.

```{r}
class(chisq.test(survey$fold, survey$clap))
names(chisq.test(survey$fold, survey$clap))
p.cs <- chisq.test(survey$fold, survey$clap)$p.value
```

- A warning message was generated  because the expected count is less than 5 in some of the cells in the table. Get the expected counts from the object returned by `chisq.test`. Hint: use the component `expected` of the object.

```{r}
res <- chisq.test(survey$fold, survey$clap)
res$expected
```


- Alternatively, an exact p-value can be obtained using a Fisher's exact test. Use `fisher.test` to test association between `fold` and `clap`. Check the syntax in the help file of the function. Check the class of the object generated and its names. Then save the p-value computed as a separate variable.

```{r}
fisher.test(survey$fold, survey$clap)
class(fisher.test(survey$fold, survey$clap))
names(fisher.test(survey$fold, survey$clap))
p.fis <- fisher.test(survey$fold, survey$clap)$p.value
```

- Print on the console all p-values computed above, rounded to 4 decimal places.

```{r}
round(p.cs, 4)
round(p.fis, 4)
```

# Tests for count tables (continued)

- Make a table of `gender` and `exercise` in the `survey` data and save the resulting object.

```{r}
t.ge <- table(survey$gender, survey$exercise)
t.ge
```

- Test for the association between `gender` and `exercise` in the `survey` data using a chi-square test. Use the table of the two variables in the function call, instead of the variables themselves. Save both the entire test result as the p-value as separate objects.

```{r}
chi.ge <- chisq.test(t.ge)
chi.ge
# Compare the result with that using the variables directly
chisq.test(survey$gender, survey$exercise)
p.ge <- chi.ge$p.value
# This is the same as
p.ge <- chisq.test(t.ge)$p.value
```

- Check the names of components in the test object. Extract the expected count table from the object to see if any of the entries are less than 5.

```{r}
names(chi.ge)
chi.ge$expected
```

- The test object also involves a component called `observed`. Check what it contains and compare it to the observed data table you saved.

```{r}
chi.ge$observed
t.ge
# The component observed from the test object corresponds to the cross-table of the variables used
```

- Compute the p-value for the test again by using simulation, instead of the chi-square approximation. This can be done by means of the `simulate.p.value` slot  in the function call. Save the computed p-value as a separate variable. 

```{r}
chisq.test(t.ge, simulate.p.value=TRUE)
p.sim <- chisq.test(t.ge, simulate.p.value=TRUE)$p.value
```


- Print both computed p-values on the console with 2 decimal places.

```{r}
round(p.ge, 2)
round(p.sim, 2)
# Note that the simulated p-value may vary every time it is run
```

