---
title: "Advanced factors (tasks)"
date: "`r format(Sys.time(), '%d %B %Y')`"
---

```{r setup, include=FALSE}
source( "../R_course_setup.R" );
```

```{r echo=FALSE}
pulse <- read.delim("../data/pulse.txt")
survey <- read.delim("../data/survey.txt")
```


# Changing levels {#levels}

In the `survey` data

- Reorder the levels of the `smokes` variable in a more natural way.

```{r}
levels(survey$smokes)
survey$smokes <- factor(survey$smokes, levels=c("never", "occas", "regul", "heavy"))
levels(survey$smokes)
```

- Make a new version of the `smokes` variable that puts the heavy, regular and occasional smokers together in the same category.

```{r}
survey$smokes2 <- survey$smokes
levels(survey$smokes2) <- c('no', 'yes', 'yes', 'yes')
```

- Tabulate the old and new smokes variables against each other to check that the category grouping worked as expected.

```{r}
table(survey$smokes, survey$smokes2)
```


# Creating factors from existing variables

Use the `survey` data for the exercises below.

- Make a categorical variable that indicates whether subjects have pulse above or at most equal to the median pulse. Beware of missing values!

```{r}
med <- median(survey$pulse, na.rm=TRUE)
# This can be done in two ways. 
# First, using a logical variable, where pulse is compared to its median
survey$pulse2 <- survey$pulse > med
# The second way is to use cut. Here, when computing minimum and maximum,
# you need to indicate that NA entries are to be ignored, as with the function to compute the median
survey$pulse2 <- cut(survey$pulse, c(min(survey$pulse, na.rm = TRUE), 
                                     med, max(survey$pulse, na.rm = TRUE)))
```

- (*) Make a new factor with four categories: `lefthanded woman`, `righthanded woman`, `lefthanded man`, `righthanded man`. Hint: use `:` to combine `gender` and `hand`.

```{r}
survey$genderhand <- survey$gender:survey$hand
```

- (*) Give the factor the levels given above: `lefthanded woman`, etc.
```{r}
levels(survey$genderhand) <- c('lefthanded woman', 'righthanded woman', 'lefthanded man', 'righthanded man')
```


# Regression with factors {#anova}

- In the `survey` data, use ANOVA to see if `pulse` is different between the `fold` groups. Hint: use `lm` to fit a model where `pulse` is explained by `fold`, then use ANOVA on the model fit.

```{r}
fit <- lm(pulse ~ fold, data=survey)
summary(fit)
anova(fit)
```

- Which category of `fold` corresponds to the intercept in this regression model? Why is that?

```{r}
# The intercept involves the first level of the factor fold. This is given below:
levels(survey$fold)[1]
# You can check this by verifying that this is the only level that does not appear explicitly in the coefficients table of the model fit summary:
summary(fit)$coefficients
```


- Reorder the levels of `fold` so that the category `fit` is the first level, and refit the model. 

```{r}
survey$fold2 <- factor(survey$fold, levels=c('right', 'neither', 'left'))
fit <- lm(pulse ~ fold2, data=survey)
summary(fit)
```

- Make boxplots of `pulse` values for each of the groups defined by `fold`. Hint: use as input for the boxplot the same formula used for the `lm`.

```{r}
boxplot(pulse ~ fold, data=survey)
# Note that the plot function would have given the same result here:
# plot(pulse ~ fold, data=survey)
```

- Fit a model to explain `pulse` by `fold`, and save the model fit in an object.


```{r}
fit1 <- lm(pulse ~ fold , data=survey)
```

- Fit now a model to explain `pulse` by both `fold` and `clap`, and save the model fit in an object. 

```{r}
fit2 <- lm(pulse ~ fold + clap, data=survey)
```

- Does `clap` have (statistical) added value in the model that already contains `fold`? Hint: use `anova` to compare the model fits already obtained.

```{r}
anova(fit1, fit2)
```


- (*) Now fit a model where `pulse` is explained by  `clap` only, and compare the fit with that using both `clap` and `fold`.  

```{r}
fit3 <- lm(pulse ~ clap , data=survey)
anova(fit3, fit2)
```


# (*) Logistic regression

In the quine data (from the library MASS, so remember to load this library using `library(MASS)`),  learning speed is studied in context of a number of variables.

- Make a logistic regression model to predict learning speed (`Lrn`) based on `Eth`, `Age` and `Sex`.

```{r}
library(MASS)
fit <- glm(Lrn ~ Eth + Age + Sex, data=quine, family=binomial)
```

- Is there a significant relationship between `Age` and `Lrn`?
  
```{r}
fit0 <- glm(Lrn ~ Eth + Sex, data=quine, family=binomial)
anova(fit, fit0, test='LRT')
```

